---
title: "Dates and character strings -- exercise solutions"
author: "Michael C Sachs"
subtitle: "Day 3, C"
---

## Read in the data

From the [data](../data/index.html), download and read in the file "med-2005-ex.rds". 

```{r}
library(here)
med2005 <- readRDS(here("data", "med-2005-ex.rds"))
```

## Dates

1. Use the `lubridate` package (try the function `wday`) to calculate the day of the week when the dispensation occurred. Do dispensations occur less frequently on weekends? 
2. Calculate the month of the dispensation. Is there a seasonal trend on the number of dispensations? 

```{r}
library(lubridate)

med2005$day_of_week <- wday(med2005$date, label = TRUE)

table(wday(med2005$date, label = TRUE)) |> plot()
table(wday(med2005$date, label = FALSE))

table(med2005$day_of_week %in% c("Sun", "Sat"))

binom.test(sum(med2005$day_of_week %in% c("Sat", "Sun")), nrow(med2005), 
           p = 2 / 7)

table(month(med2005$date, label = TRUE)) |> plot()
```

3. Use `dplyr` or `data.table` to create a new variable that is the last dispensation for each individual during the year (using `group_by` then `mutate` or `:=` with `by`). Then calculate the number of days between each dispensation and the last one. What is the average number of days? 

::: {.panel-tabset}

## `data.table` 

```{r}
library(data.table)

med2005dt <- data.table(med2005)

med2005dt[, last_date := max(date), 
          by = .(pid)]


med2005dt[, num_days := as.numeric(date - last_date)]
med2005dt[num_days == 0, num_days := NA]

summary(med2005dt)
hist(med2005dt$num_days)
```


## `dplyr` 

```{r}
library(dplyr)

med2005b <- med2005 |> group_by(pid) |> 
  mutate(last_date = max(date))

med2005b <- med2005b |> mutate(num_days = as.numeric(date - last_date))
med2005b$num_days[med2005b$num_days == 0] <- NA

summary(med2005b)
hist(med2005b$num_days)
```
:::

## Regular expressions


The "atc" variable contains the ATC code for the drug that was dispensed. This is a standardized classification system for drugs, see [https://www.whocc.no/atc/structure_and_principles/](https://www.whocc.no/atc/structure_and_principles/). 

1. Count how many dispensations there were for drugs in the class "J01". Use the `grepl` function with a pattern. 2. Are there any "invalid" ATC codes in the dataset? 

```{r}

grepl("^J01",med2005$atc) |>head()
grepl("^J01.{4}$",med2005$atc) |> table()

## replace
gsub("^J01.{4}$", "mydrug", med2005$atc) |>head()

## check for invalid atcs
grepl("^[A-Z]{1}[0-9]{2}[A-Z]{2}[0-9]{2}$", 
      c(med2005$atc[1:5], "100", "J01DD09b")) 

## no invalid atcs
grepl("^[A-Z]{1}[0-9]{2}[A-Z]{2}[0-9]{2}$", 
  med2005$atc) |> table()
```

