---
title: "Tidying and reshaping -- exercise solutions"
author: "Michael C Sachs"
subtitle: "Day 3, B"
editor_options: 
  chunk_output_type: console
---


## Tidying our mean sd function

1. Write a tidy method for our `mean_sd` function and try it out on the penguins dataset. 

```{r}
library(palmerpenguins)
library(broom)

#Write a tidy method for our mean_sd function and try it out on the penguins dataset.

mean_sd <- function(x, na.rm = TRUE) {
  
  xname <- deparse1(substitute(x))  # gets the name of the variable
  
  res <- c(mean = mean(x, na.rm = na.rm), 
           sd = sd(x, na.rm = na.rm))
  
  class(res) <- "meansd"
  attr(res, "variable") <- xname
  attr(res, "sampsize") <- length(x)
  
  res
  
}

print.meansd <- function(x, digits = 2) {
  
  cat(attr(x, "variable"), ": ",
      paste0(
        round(x["mean"], digits = digits), 
        " (", 
        round(x["sd"], digits = digits), 
        ")\n"))
  
}

tidy.meansd <- function(x) {
  
  data.frame(variable = attr(x, "variable"), 
             mean = x["mean"], 
             sd = x["sd"], 
             sample.size = attr(x, "sampsize")
             )
  
}


mean_sd(penguins$body_mass_g)
tidy(mean_sd(penguins$body_mass_g))

```


2. Apply the `mean_sd` function to the penguins body mass in grams by species and sex. Organize the results into a table suitable for publication, where it is easy to compare the two sexes.

::: {.panel-tabset}

## Base R

```{r}

reslist <- split(penguins$body_mass_g, list(penguins$species, penguins$sex)) |>
  lapply(FUN = \(x) tidy(mean_sd(x))) 

resdf <- cbind(do.call(rbind, reslist), 
      do.call(rbind, names(reslist)|> strsplit(split="\\.")))
colnames(resdf)[5:6] <- c("species", "sex")

reshape(resdf[, -1], direction = "wide", idvar = "species", timevar = "sex")
```

## `data.table` 

```{r}
library(data.table)

pengdt <- data.table(penguins)
peng_summary <- pengdt[, (tidy(mean_sd(body_mass_g))), 
                       by = list(species, sex)]

dcast(peng_summary[!is.na(sex)], species + variable ~ sex, 
      value.var = c("mean", "sd", "sample.size"))

```


## `dplyr`


```{r}
library(dplyr)
library(tidyr)
penguins |> group_by(species, sex) |>
  summarize(tidy(mean_sd(body_mass_g))) |> 
  filter(!is.na(sex)) |> 
  pivot_wider(names_from = "sex", 
              values_from = c("mean", "sd", "sample.size"))

```

:::

## Tidying the national patient register dataset

Load the LPR data example from `"https://sachsmc.github.io/r-programming/data/lpr-ex.rds"`

```{r}
library(here)
lpr <- readRDS(here("data", "lpr-ex.rds"))
```

1. Reshape the data into wide, where the columns are the primary diagnosis (hdia) at each visit number
2. Reshape the data into longer format, where all of the diagnoses are stored in a single variable, with another variable indicating the primary diagnosis. 
3. Create a new variable for each participant which equals TRUE if they had any diagnosis of either D150, D152, or D159 before the date 1 January 2010.


::: {.panel-tabset}
## base R

```{r}
## 1. 
lprwide <- subset(lpr, select= c(pid, age, sex, visit, hdia)) |> ## drop the other diag columns and date
  reshape(direction = "wide", 
        idvar = "pid", timevar = "visit", 
        v.names = c("hdia"))
head(lprwide)

## 2. 
lprtmp <- lpr
colnames(lprtmp)[which(colnames(lprtmp) == "hdia")] <- "diag0"
lprlong <- reshape(lprtmp, direction = "long", 
        varying = grep("^diag", names(lprtmp)), 
        sep = "", timevar = "diagnum")
lprlong$maindiag <- lprlong$diagnum == 0
lprlong <- subset(lprlong[order(lprlong$pid, lprlong$visit),], 
                  !is.na(diag))


head(lprlong)

## 3. 

lprlong <- split(lprlong, lprlong$pid) |> 
  lapply(FUN = \(df) {
    
    chk <- any(df$diag[df$indat <= as.Date("2010-01-01")] %in% c("D150", "D152", "D159"))
    if(length(chk) == 0) chk <- FALSE
    df$ddiag_pre2010 <- chk
    df
  }) |> 
  unsplit(lprlong$pid)

summary(lprlong)
```


## `data.table`

```{r}
## 1.
lprdt <- data.table(lpr)
dcast(lprdt[, .(pid, indat, visit, hdia)], 
      pid ~ visit, value.var = "hdia")

# 2. 
lprlong <- melt(lprdt, id.vars = c("pid", "age", "sex", "indat", "visit"), 
                variable.name = "diagtype", value.name = "diag")
lprlong <- lprlong[!is.na(diag)]
lprlong[, main_diag := diagtype == "hdia"]

lprlong
setorder(lprlong, pid, indat)
lprlong

# 3. 


lprlong <- lprlong[, ddiag_pre2010 := any(diag[indat <= as.Date("2010-01-01")] %in% 
          c("D150", "D152", "D159")), 
        by = .(pid)]

summary(lprlong)
lprdt[pid == "A048"]
```


## tidyverse

```{r}
lprtbl <- as_tibble(lpr)

## 1. 
lprtbl |> select(pid, age, sex, visit, hdia) |> 
  pivot_wider(names_from = visit, values_from = hdia, names_prefix = "visit")

## 2. 

lprlong <- lprtbl |> pivot_longer(cols = hdia:diag6, values_to = "diag") |> 
  mutate(maindiag = name == "hdia") |> 
  filter(!is.na(diag))
lprlong

## 3. 
lprlong <- lprlong |> group_by(pid) |> 
  mutate(ddiag_pre2010 = any(diag[indat <= as.Date("2010-01-01")] %in% c("D150", "D152", "D159")))

summary(lprlong)
```
:::


## Merging and manipulation

The objective of this exercise is to describe the distribution of the number of days between hospitalizations and drug dispensations by age and sex, among people who get drugs dispensed following a hospitalization. 

1. Import and merge the drug register data with the long version of the hospitalization register from the previous section. What type of join is needed here? 
2. Create a new variable that counts the number of drug dispensations in the 3 months following a hospitalization. 
3. Summarize the variable by age and sex. Try making a graphical summary.


Reading in the data using lapply

```{r}
medsl <- lapply(2005:2010, \(y) {
  readRDS(here("data", paste0("med-", y, "-ex.rds")))
})
```

or a loop

```{r}
medsl <- vector(mode = "list", length = 6)
names(medsl) <- 2005:2010
for(i in 1:length(medsl)) {
  medsl[[i]] <- readRDS(here("data", paste0("med-", names(medsl)[i], "-ex.rds")))
}

meds <- do.call(rbind, medsl)
```

```{r}
library(lubridate)
```

::: {.panel-tabset}
## base R

```{r}
lprmed <- merge(subset(lprlong, maindiag == TRUE), meds, by = "pid")
lprmed$days_to_dispensation <- with(lprmed, date - indat)
lprmed <- subset(lprmed, days_to_dispensation >= 0 & 
                   days_to_dispensation <= dmonths(3))

lprmed2 <- do.call("rbind", split(lprmed, list(lprmed$pid, lprmed$visit), drop = TRUE) |> 
  lapply(\(df) {
    df$num_drugs_3m <- nrow(df)
    unique(df[, c("pid", "age", "sex", "indat", "visit", "diag", "num_drugs_3m")])
  }))

lprmed2 <- lprmed2[order(lprmed2$pid, lprmed2$visit),]

lprmed2$agecat <- cut(lprmed2$age, c(25, 40, 60, 80, 100))

by(lprmed2$num_drugs_3m, list(lprmed2$sex, 
                              lprmed2$agecat), 
   table)

par(mfrow = c(2, 1)) 
for(s in c("m", "f")) {
  
  boxplot(num_drugs_3m ~ agecat, data = subset(lprmed2, sex == s), 
       main = paste0("Sex: ", s))
  
}
par(mfrow = c(1,1))
```


## dplyr

```{r}
lprmeddply <- lprlong |> filter(maindiag == TRUE) |> 
  mutate(indat_plus_3m = indat + dmonths(3)) |> 
  inner_join(meds, join_by(pid, 
                           between(y$date, x$indat, x$indat_plus_3m))) |>
  group_by(pid, visit) |> 
  summarize(sex = sex[1], age = age[1], agecat =  cut(age[1], c(25, 40, 60, 80, 100)),
              n_drugs_3m = n())
  
lprmeddply |> group_by(sex, agecat, n_drugs_3m) |> 
  summarize(count_ndrugs = n())

with(lprmeddply, by(n_drugs_3m, list(sex, agecat), table))
```


## data.table

```{r}
lprdt <- data.table(lprlong)[maindiag == TRUE]
meds <- data.table(meds)
meds[, drugdate := date]
lprdt[, indate_plus_3m := indat + dmonths(3)]

lprmeds <- meds[lprdt, , on = .(pid == pid, date >= indat), nomatch = NULL][drugdate <= indate_plus_3m]
lprmeds2 <- lprmeds[, .(sex = sex[1], age = age[1], agecat = cut(age[1], c(25, 40, 60, 80, 100)), 
            n_drugs_3m = .N), by = .(pid, visit)]

with(lprmeds2, by(n_drugs_3m, list(sex, agecat), table))
```

:::
